ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-05-09 notebooks/moons.qmd
more results

```{julia}
λ₁ = 0.25
λ₂ = 0.75
λ₃ = 0.75
Λ = [λ₁, λ₂, λ₃]

# Benchmark generators:
generator_dict = Dict(
    "Wachter" => WachterGenerator(λ=λ₁),
    "REVISE" => REVISEGenerator(λ=λ₁),
    "Schut" => GreedyGenerator(),
    "ECCCo" => ECCCoGenerator(λ=Λ),
)
```

@@

ces = Dict{Any,Any}()
plts = []
for (name, generator) in generator_dict
    ce = generate_counterfactual(
        x_factual, target, counterfactual_data, M, generator;
        initialization=:identity, 
        converge_when=:generator_conditions,
    )
    plt = Plots.plot(ce, title=name, alpha=0.2, cbar=false, axis=nothing)
    if name == "ECCCo"
        _X = distance_from_energy(ce, return_conditionals=true)
        Plots.scatter!(
            _X[1,:],_X[2,:], color=:purple, shape=:star5, 
            ms=10, label="x̂|$target", alpha=0.5
        )
    end
    push!(plts, plt)
    ces[name] = ce
end
plt = Plots.plot(plts..., size=(500,520))
display(plt)
savefig(plt, joinpath(output_images_path, "moons_poc.png"))
```

```{julia}
evaluate(collect(values(ces)))
```

@@

bmk = benchmark(
    counterfactual_data; 
    models=model_dict, 
    generators=generator_dict, 
    measure=measures,
    suppress_training=true, dataname="Moons",
    n_individuals=5,
    target=1, factual=0,
    initialization=:identity,
    converge_when=:generator_conditions,
)
CSV.write(joinpath(output_path, "moons_benchmark.csv"), bmk())
```

```{julia}
@chain bmk() begin
    @group_by(dataname, generator, model, variable)
    @summarize(mean=mean(value),sd=std(value))
    @ungroup
    @filter(variable == "distance_from_energy")
end
```

>>>>

```{julia}
λ₁ = 0.1
λ₂ = 0.25
λ₃ = 0.25
Λ = [λ₁, λ₂, λ₃]

# Benchmark generators:
generator_dict = Dict(
    "Wachter" => WachterGenerator(λ=λ₁),
    "REVISE" => REVISEGenerator(λ=λ₁),
    "Schut" => GreedyGenerator(),
    "ECCCo" => ECCCoGenerator(λ=Λ),
    "ECCCo (no CP)" => ECCCoGenerator(λ=[λ₁, 0.0, λ₃]),
    "ECCCo (no EBM)" => ECCCoGenerator(λ=[λ₁, λ₂, 0.0]),
)
```

@@

ces = Dict{Any,Any}()
plts = []
for (name, generator) in generator_dict
    ce = generate_counterfactual(
        x_factual, target, counterfactual_data, M, generator;
        initialization=:identity, 
        converge_when=:generator_conditions,
    )
    plt = Plots.plot(ce, title=name, alpha=0.2, cbar=false, axis=nothing)
    if contains(name, "ECCCo")
        _X = distance_from_energy(ce, return_conditionals=true)
        Plots.scatter!(
            _X[1,:],_X[2,:], color=:purple, shape=:star5, 
            ms=10, label="x̂|$target", alpha=0.5
        )
    end
    push!(plts, plt)
    ces[name] = ce
end
plt = Plots.plot(plts..., size=(650,500))
display(plt)
savefig(plt, joinpath(output_images_path, "moons_poc.png"))
```

@@

bmks = []
for target in sort(unique(labels))
    for factual in sort(unique(labels))
        if factual == target
            continue
        end
        bmk = benchmark(
            counterfactual_data; 
            models=model_dict, 
            generators=generator_dict, 
            measure=measures,
            suppress_training=true, dataname="Moons",
            n_individuals=25,
            target=target, factual=factual,
            initialization=:identity,
            converge_when=:generator_conditions,
        )
        push!(bmks, bmk)
    end
end
bmk = reduce(vcat, bmks)
CSV.write(joinpath(output_path, "moons_benchmark.csv"), bmk())
```

### 2023-05-09 notebooks/poc.qmd
work on table
more results

```{julia}
include("notebooks/setup.jl")
eval(setup_notebooks)
```

```{julia}
# Counteractual data:
counterfactual_data = load_linearly_separable(1000)
X = counterfactual_data.X
y = counterfactual_data.y
labels = counterfactual_data.output_encoder.labels
input_dim, nobs = size(X)
batch_size = Int(round(nobs/10))
epochs = 100
```

@@

λ₁ = 0.1
λ₂ = 0.5
λ₃ = 0.5
Λ = [λ₁, λ₂, λ₃]

@@

generator_dict = OrderedDict(
    "Wachter (γ=$γ)" => WachterGenerator(λ = λ₁),
    "Schut" => GreedyGenerator(λ = λ₁),
    "REVISE" => REVISEGenerator(λ = λ₁),
    "ECCCo" => ECCCoGenerator(λ = Λ),
)

ces = Dict{Any,Any}()
plts = []
for (name, generator) in generator_dict
    ce = generate_counterfactual(
        x_factual, target, counterfactual_data, M, generator;
        initialization=:identity, 
        converge_when=:generator_conditions,
        decision_threshold=γ,
    )
    plt = Plots.plot(ce, title=name, alpha=0.2, cbar=false, axis=nothing, length_out=10)
    if name == "ECCCo"
        _X = distance_from_energy(ce, return_conditionals=true)
        Plots.scatter!(
            _X[1,:],_X[2,:], color=:purple, shape=:star5, 
            ms=10, label="x̂|$target", alpha=0.1
        )
    end
    push!(plts, plt)
    ces[name] = ce
end
plt = Plots.plot(plts..., size=(500,520))
display(plt)
savefig(plt, joinpath(output_images_path, "poc.png"))
```

@@

scale = 0.1
for (x, y) in inputs
    push!(u, scale * gradient(f2, [x, y][:,:])[1][1])
    push!(v, scale * gradient(f2, [x, y][:,:])[1][2])
end

>>>>

```{julia}
include("$(pwd())/notebooks/setup.jl")
eval(setup_notebooks)
```

```{julia}
# Counteractual data:
n_obs = 1000
counterfactual_data = load_blobs(n_obs; cluster_std=0.1, center_box=(-1. => 1.))
X = counterfactual_data.X
y = counterfactual_data.y
labels = counterfactual_data.output_encoder.labels
input_dim, nobs = size(X)
batch_size = Int(round(nobs/10))
epochs = 100
```

@@

λ₁ = 0.1
λ₂ = 0.4
λ₃ = 0.4
Λ = [λ₁, λ₂, λ₃]

@@

opt = Flux.Optimise.Descent(0.01)

generator_dict = OrderedDict(
    "Wachter" => WachterGenerator(λ = λ₁, opt=opt),
    "Schut" => GreedyGenerator(λ = λ₁),
    "REVISE" => REVISEGenerator(λ = λ₁, opt=opt),
    "ECCCo" => ECCCoGenerator(λ = Λ, opt=opt),
)

ces = Dict{Any,Any}()
plts = []
for (name, generator) in generator_dict
    ce = generate_counterfactual(
        x_factual, target, counterfactual_data, M, generator;
        initialization=:identity, 
        converge_when=:generator_conditions,
    )
    plt = Plots.plot(
        ce, title=name, alpha=0.2, cbar=false, 
        axis=nothing, length_out=10, contour_alpha=1.0,
    )
    if name == "ECCCo"
        _X = distance_from_energy(ce, return_conditionals=true)
        Plots.scatter!(
            _X[1,:],_X[2,:], color=:purple, shape=:star5, 
            ms=10, label="x̂|$target", alpha=0.1
        )
    end
    push!(plts, plt)
    ces[name] = ce
end
plt = Plots.plot(plts..., size=(500,520))
display(plt)
savefig(plt, joinpath(output_images_path, "poc.png"))
```

@@

scale = 0.1
for (x, y) in inputs
    push!(u, scale * gradient(f1, [x, y][:,:])[1][1])
    push!(v, scale * gradient(f1, [x, y][:,:])[1][2])
end

### 2023-05-09 notebooks/tables.Rmd
more results

```{r}
files <- list.files("artifacts/results/")
dt <- lapply(files[grepl("_benchmark.csv", files)], function(x) {
    fread(file.path("artifacts/results/", x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
dt[,ce:=NULL]
synth <- c("Moons", "Circles")
dt[,source:=ifelse(dataname %in% synth, "synthetic", "real")]
```

@@

```{r}
measures <- c("distance_from_energy", "distance_from_targets")
tab <- dt[
    variable %in% measures & source=="synthetic",
    .(
      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
      val = mean(value)
    ),
    .(dataname, generator, model, variable)
]
tab[,min_val:=val==min(val),.(model, dataname, variable)]
tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
tab[,val:=NULL]
tab[,min_val:=NULL]
tab[generator == "eccco", generator:="ECCCo"]
tab[generator == "revise", generator:="REVISE"]
tab[generator == "generic", generator:="Wachter"]
tab[generator == "wachter", generator:="Wachter"]
tab[generator == "greedy", generator:="Schut"]
n_data <- length(unique(tab$dataname))
tab <- dcast(tab, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(c("Non-conformity ↓","Implausibility ↓"),n_data)
)
kbl(
  tab, caption = "Results for synthetic datasets. Standard deviation in parentheses. \\label{tab:results-synth}", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = c("", "", "", "\\addlinespace") 
) %>%
  # kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(c(" " = 2, "Moons" = 2)) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable("paper/contents/table_synth.tex")
```

>>>>

```{r}
files <- list.files("artifacts/results/")
dt <- lapply(files[grepl("_benchmark.csv", files)], function(x) {
    fread(file.path("artifacts/results/", x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
dt[,ce:=NULL]
synth <- c("Moons", "Circles", "Linearly Separable")
dt[,source:=ifelse(dataname %in% synth, "synthetic", "real")]
```

@@

```{r}
measures <- c("distance_from_energy", "distance_from_targets")
tab <- dt[
    variable %in% measures & source=="synthetic",
    .(
      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
      val = mean(value),
      std = sd(value)
    ),
    .(dataname, generator, model, variable)
]
tab[,min_val:=val==min(val),.(model, dataname, variable)]
tab[,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[one_std_wachter==T,value:=paste0(value,"*")]
tab[two_std_wachter==T,value:=paste0(value,"*")]
tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
tab[,val:=NULL]
tab[,std:=NULL]
tab[,min_val:=NULL]
tab[,two_std_wachter:=NULL]
tab[,one_std_wachter:=NULL]
n_data <- length(unique(tab$dataname))
tab <- dcast(tab, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(c("Non-conformity ↓","Implausibility ↓"),n_data)
)
kbl(
  tab, caption = "Results for synthetic datasets. Standard deviations across samples are shown in parentheses. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-synth} \\newline", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = c("", "", "", "\\addlinespace") 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(c(" " = 2, "Circles" = 2, "Linearly Separable" = 2, "Moons" = 2)) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable("paper/contents/table_synth.tex")
```

### 2023-05-09 paper/paper.tex
more results

\begin{itemize}
  \item BatchNorm does not seem compatible with JEM
  \item Coverage and temperature impacts CCE in somewhat unpredictable ways
  \item It seems that models that are not explicitly trained for generative task, still learn it implictly
  \item Batch size seems to impact quality of generated samples (at inference, but not so much during JEM training)
  \item ECCCo is sensitive to optimizer (Adam works well), learning rate and distance metric (l1 currently only one that works)
  \item SGLD takes time 
  \item REVISE has benefit of lower dimensional space
  \item For MNIST it seems that ECCCo is better at reducing pixel values than increasing them (better at erasing than writing)
  \item JEMs are more difficult to train
\end{itemize}

>>>>

\begin{itemize}
  \item BatchNorm does not seem compatible with JEM
  \item Coverage and temperature impacts CCE in somewhat unpredictable ways
  \item It seems that models that are not explicitly trained for generative task, still learn it implictly
  \item Batch size seems to impact quality of generated samples (at inference, but not so much during JEM training)
  \item SGLD takes time 
  \item REVISE has benefit of lower dimensional space
  \item For MNIST it seems that ECCCo is better at reducing pixel values than increasing them (better at erasing than writing)
  \item JEMs are more difficult to train
  \item There is a tradeoff: higher cost vs. higher conformity/plausibility
  \item Results are sensitive to choices of penalty strength and step size
  \item Counterfactuals may end up looking fairly homogenous
  \item For MNIST data we found CP to have little effect
  \item JEMs themselves are sensitive to scale
  \item ECCCo can backfire, in case generative property of model is poor
\end{itemize}
